\vspace{30mm} {\Large\bf RISC-V Duo-PLIC for the Advancced Interrupt Architecture} \\ {\large Document Version \DuoPLICRev}

### 2023-05-09 doc/src/MSLevel.tex
Change descriptions of high-half CSRs to match latest Priv ISA
Signed-off-by: John Hauser <31252952+jhauser-us@users.noreply.github.com>

\begin{commentary}
While many of the standard registers such as\/ \z{mip}/\z{miph}
and\/ \z{mie}/\z{mieh} have space for major interrupts
only in the range 0--63, custom interrupts with numbers
64 and above are conceivable with added custom support.
CSRs\/ \z{mtopi} (Section~\ref{sec:mtopi})
and\/ \z{stopi} (Section~\ref{sec:stopi}) allow for
major interrupt numbers potentially as large as 4095.
\end{commentary}

@@

For whichever standard local interrupts are
implemented, the corresponding bits in CSRs
\z{mip}/\z{miph} and \z{mie}/\z{mieh} must be writable, and the
corresponding bits in \z{mideleg}/\z{midelegh} (if those CSRs exist
because supervisor mode is implemented) must each either be writable or
be hardwired to zero.
An occurrence of a local interrupt event causes the interrupt-pending bit
in \z{mip}/\z{miph} to be set to one.
This bit then remains set until cleared by software.

As established by the {\RISCV} Privileged Architecture, an interrupt
traps to \mbox{M-mode} whenever all of the following are true:
(a)~either the current privilege mode is \mbox{M-mode} and
machine-level interrupts are enabled by the MIE bit of \z{mstatus}, or
the current privilege mode has less privilege than \mbox{M-mode};
(b)~matching bits in \z{mip}/\z{miph} and \z{mie}/\z{mieh} are both
one; and
(c)~if \z{mideleg} exists, the corresponding bit in
\z{mideleg}/\z{midelegh} is zero.

@@

Each byte of a valid \z{iprio}$k$ register is either a read-only zero
or a {\WARL} unsigned integer field implementing exactly IPRIOLEN bits.
For a given interrupt number, if the corresponding bit in
\z{mie}/\z{mieh} is read-only zero, then the interrupt's priority
number in the \z{iprio} array must be read-only zero as well.
The priority number for a machine-level external interrupt (bits 31:24
of register \z{iprio2}) must also be read-only zero.
Aside from these two restrictions, implementations may freely choose
which priority number fields are settable and which are read-only
zeros.
If all bytes in the \z{iprio} array are read-only zeros, priorities
can be configured only for external interrupts, not for any other
interrupts.

@@

The value of \z{mtopi} is zero unless there is an interrupt pending in
\z{mip}/\z{miph} and enabled in \z{mie}/\z{mieh} that is not delegated
to a lower privilege level.
When there is a pending-and-enabled major interrupt for machine
level, field IID (Interrupt Identity) is the major identity number of
the highest-priority interrupt, and field IPRIO indicates its priority.

@@

When supervisor mode is implemented, the Advanced Interrupt
Architecture adds a facility for software filtering
of interrupts and for virtual interrupts, making use of new
CSRs \z{mvien}/\z{mvienh} (Machine Virtual Interrupt Enables) and
\z{mvip}/\z{mviph} (Machine Virtual Interrupt-Pending bits).
\emph{Interrupt filtering} permits a supervisor-level
interrupt (SEI or SSI) or local or custom interrupt
to trap to \mbox{M-mode} and then be selectively delegated by
software to supervisor level, even while the corresponding bit in
\z{mideleg}/\z{midelegh} remains zero.
The same hardware may also, under the right circumstances, allow
machine level to assert \emph{virtual interrupts} to supervisor level
that have no connection to any real interrupt events.

Just as with CSRs \z{mip}/\z{miph}, \z{mie}/\z{mieh}, and
\z{mideleg}/\z{midelegh}, each bit of registers \z{mvien}/\z{mvienh}
and \z{mvip}/\z{mviph} corresponds with an interrupt number in the
range 0--63.
When a bit in \z{mideleg}/\z{midelegh} is zero and the matching
bit in \z{mvien}/\z{mvienh} is one, then the same bit position
in \z{sip}/\z{siph} is an alias for the corresponding bit in
\z{mvip}/\z{mviph}.
A bit in \z{sip}/\z{siph} is read-only zero when the corresponding bits
in \z{mideleg}/\z{midelegh} and \z{mvien}/\z{mvienh} are both zero.
The combined effects of \z{mideleg}/\z{midelegh} and
\z{mvien}/\z{mvienh} on \z{sip}/\z{siph} and \z{sie}/\z{sieh} are
summarized in Table~\ref{tab:intrFilteringForS}.

@@

A bit in \z{sie}/\z{sieh} is writable if and only if the corresponding
bit is set in either \z{mideleg}/\z{midelegh} or \z{mvien}/\z{mvienh}.
When an interrupt is delegated by \z{mideleg}/\z{midelegh}, the
writable bit in \z{sie}/\z{sieh} is an alias of the corresponding bit
in \z{mie}/\z{mieh};
else it is an independent writable bit.
As usual, bits that are not writable in \z{sie}/\z{sieh} must be
read-only zeros.

If a bit of \z{mideleg}/\z{midelegh} is zero and the corresponding bit
in \z{mvien}/\z{mvienh} is changed from zero to one, then the value
of the matching bit in \z{sie}/\z{sieh} becomes {\unspecified}.
Likewise, if a bit of \z{mvien}/\z{mvienh} is one and the corresponding
bit in \z{mideleg}/\z{midelegh} is changed from one to zero, the value
of the matching bit in \z{sie}/\z{sieh} again becomes {\unspecified}.

For interrupt numbers 13--63, implementations may freely choose which
bits of \z{mvien}/\z{mvienh} are writable and which bits are read-only
zero or one.
If such a bit in \z{mvien}/\z{mvienh} is read-only zero (preventing the
virtual interrupt from being enabled), the same bit should be read-only
zero in \z{mvip}/\z{mviph}.
All other bits for interrupts 13--63 must be writable in
\z{mvip}/\z{mviph}.

@@

When the hypervisor extension is implemented, if a bit is
zero in the same position in both \z{mideleg}/\z{midelegh}
and \z{mvien}/\z{mvienh}, then that bit is read-only zero in
\z{hideleg}/\z{hidelegh} (in addition to being read-only zero
in \z{sip}/\z{siph}, \z{sie}/\z{sieh}, \z{hip}, and \z{hie}).
But if a bit for one of interrupts 13--63 is a one in
either \z{mideleg}/\z{midelegh} or \z{mvien}/\z{mvienh},
then the same bit in \z{hideleg}/\z{hidelegh} may be writable
or may be read-only zero, depending on the implementation.
No bits in \z{hideleg}/\z{hidelegh} are ever read-only ones.
The {\RISCV} Privileged Architecture further
constrains bits 12:0 of \z{hideleg}.

When supervisor mode is implemented, the minimal required
implementation of \z{mvien}/\z{mvienh} and \z{mvip}/\z{mviph} has all
bits being read-only zeros except for \z{mvip}
bits 1 and~9, and sometimes bit~5, each of which is
an alias of an existing writable bit in \z{mip}.
(Although, as noted, it is strongly recommended
that bit 9 of \z{mvien} also be writable.)
When supervisor mode is not implemented, registers \z{mvien}/\z{mvienh}
and \z{mvip}/\z{mviph} do not exist.

@@

If a standard local interrupt
becomes pending (=~1) in \z{sip}/\z{siph}, the bit in \z{sip}/\z{siph}
is writable and will remain set until cleared by software.

Just as for machine level, the taking of interrupt traps at supervisor
level remains essentially the same as specified by the {\RISCV}
Privileged Architecture.
An interrupt traps into \mbox{S-mode} (or \mbox{HS-mode}) whenever all
of the following are true:
(a)~either the current privilege mode is \mbox{S-mode} and
supervisor-level interrupts are enabled by the SIE bit of \z{sstatus},
or the current privilege mode has less privilege than \mbox{S-mode};
(b)~matching bits in \z{sip}/\z{siph} and \z{sie}/\z{sieh} are both
one, or, if the hypervisor extension is implemented, matching bits in
\z{hip} and \z{hie} are both one; and
(c)~if the hypervisor extension is implemented, the corresponding bit
in \z{hideleg}/\z{hidelegh} is zero.

@@

For a given interrupt number, if the corresponding bit in
\z{sie}/\z{sieh} is read-only zero, then the interrupt's priority
number in the supervisor-level \z{iprio} array must be read-only zero
as well.
The priority number for a supervisor-level external interrupt
(bits 15:8 of \z{iprio2}) must also be read-only zero.
Aside from these two restrictions, implementations may freely choose
which priority number fields are settable and which are read-only
zeros.

@@

The value of \z{stopi} is zero unless:
(a)~there is an interrupt that is both pending in \z{sip}/\z{siph}
and enabled in \z{sie}/\z{sieh}, or, if the hypervisor extension is
implemented, both pending in \z{hip} and enabled in \z{hie}; and
(b)~the interrupt is not delegated to a lower privilege level
(by \z{hideleg}, if the hypervisor extension is implemented).
When there is a pending-and-enabled major interrupt for supervisor
level, field IID is the major identity number of the highest-priority
interrupt, and field IPRIO indicates its priority.

>>>>

\begin{commentary}
While many of the standard registers such as\/ \z{mip}
and\/ \z{mie} have space for major interrupts
only in the range 0--63, custom interrupts with numbers
64 and above are conceivable with added custom support.
CSRs\/ \z{mtopi} (Section~\ref{sec:mtopi})
and\/ \z{stopi} (Section~\ref{sec:stopi}) allow for
major interrupt numbers potentially as large as 4095.
\end{commentary}

@@

For whichever standard local interrupts are
implemented, the corresponding bits in CSRs
\z{mip} and \z{mie} must be writable, and the
corresponding bits in \z{mideleg} (if that CSR exists
because supervisor mode is implemented) must each either be writable or
be hardwired to zero.
An occurrence of a local interrupt event causes the interrupt-pending bit
in \z{mip} to be set to one.
This bit then remains set until cleared by software.

As established by the {\RISCV} Privileged Architecture, an interrupt
traps to \mbox{M-mode} whenever all of the following are true:
(a)~either the current privilege mode is \mbox{M-mode} and
machine-level interrupts are enabled by the MIE bit of \z{mstatus}, or
the current privilege mode has less privilege than \mbox{M-mode};
(b)~matching bits in \z{mip} and \z{mie} are both
one; and
(c)~if \z{mideleg} exists, the corresponding bit in
\z{mideleg} is zero.

@@

Each byte of a valid \z{iprio}$k$ register is either a read-only zero
or a {\WARL} unsigned integer field implementing exactly IPRIOLEN bits.
For a given interrupt number, if the corresponding bit in
\z{mie} is read-only zero, then the interrupt's priority
number in the \z{iprio} array must be read-only zero as well.
The priority number for a machine-level external interrupt (bits 31:24
of register \z{iprio2}) must also be read-only zero.
Aside from these two restrictions, implementations may freely choose
which priority number fields are settable and which are read-only
zeros.
If all bytes in the \z{iprio} array are read-only zeros, priorities
can be configured only for external interrupts, not for any other
interrupts.

@@

The value of \z{mtopi} is zero unless there is an interrupt pending in
\z{mip} and enabled in \z{mie} that is not delegated
to a lower privilege level.
When there is a pending-and-enabled major interrupt for machine
level, field IID (Interrupt Identity) is the major identity number of
the highest-priority interrupt, and field IPRIO indicates its priority.

@@

When supervisor mode is implemented, the Advanced Interrupt
Architecture adds a facility for software filtering
of interrupts and for virtual interrupts, making use of new
CSRs \z{mvien} (Machine Virtual Interrupt Enables) and
\z{mvip} (Machine Virtual Interrupt-Pending bits).
\emph{Interrupt filtering} permits a supervisor-level
interrupt (SEI or SSI) or local or custom interrupt
to trap to \mbox{M-mode} and then be selectively delegated by
software to supervisor level, even while the corresponding bit in
\z{mideleg} remains zero.
The same hardware may also, under the right circumstances, allow
machine level to assert \emph{virtual interrupts} to supervisor level
that have no connection to any real interrupt events.

Just as with CSRs \z{mip}, \z{mie}, and
\z{mideleg}, each bit of registers \z{mvien}
and \z{mvip} corresponds with an interrupt number in the
range 0--63.
When a bit in \z{mideleg} is zero and the matching
bit in \z{mvien} is one, then the same bit position
in \z{sip} is an alias for the corresponding bit in
\z{mvip}.
A bit in \z{sip} is read-only zero when the corresponding bits
in \z{mideleg} and \z{mvien} are both zero.
The combined effects of \z{mideleg} and
\z{mvien} on \z{sip} and \z{sie} are
summarized in Table~\ref{tab:intrFilteringForS}.

@@

A bit in \z{sie} is writable if and only if the corresponding
bit is set in either \z{mideleg} or \z{mvien}.
When an interrupt is delegated by \z{mideleg}, the
writable bit in \z{sie} is an alias of the corresponding bit
in \z{mie};
else it is an independent writable bit.
As usual, bits that are not writable in \z{sie} must be
read-only zeros.

If a bit of \z{mideleg} is zero and the corresponding bit
in \z{mvien} is changed from zero to one, then the value
of the matching bit in \z{sie} becomes {\unspecified}.
Likewise, if a bit of \z{mvien} is one and the corresponding
bit in \z{mideleg} is changed from one to zero, the value
of the matching bit in \z{sie} again becomes {\unspecified}.

For interrupt numbers 13--63, implementations may freely choose which
bits of \z{mvien} are writable and which bits are read-only
zero or one.
If such a bit in \z{mvien} is read-only zero (preventing the
virtual interrupt from being enabled), the same bit should be read-only
zero in \z{mvip}.
All other bits for interrupts 13--63 must be writable in
\z{mvip}.

@@

When the hypervisor extension is implemented, if a bit is
zero in the same position in both \z{mideleg}
and \z{mvien}, then that bit is read-only zero in
\z{hideleg} (in addition to being read-only zero
in \z{sip}, \z{sie}, \z{hip}, and \z{hie}).
But if a bit for one of interrupts 13--63 is a one in
either \z{mideleg} or \z{mvien},
then the same bit in \z{hideleg} may be writable
or may be read-only zero, depending on the implementation.
No bits in \z{hideleg} are ever read-only ones.
The {\RISCV} Privileged Architecture further
constrains bits 12:0 of \z{hideleg}.

When supervisor mode is implemented, the minimal required
implementation of \z{mvien} and \z{mvip} has all
bits being read-only zeros except for \z{mvip}
bits 1 and~9, and sometimes bit~5, each of which is
an alias of an existing writable bit in \z{mip}.
(Although, as noted, it is strongly recommended
that bit 9 of \z{mvien} also be writable.)
When supervisor mode is not implemented, registers \z{mvien}
and \z{mvip} do not exist.

@@

If a standard local interrupt
becomes pending (=~1) in \z{sip}, the bit in \z{sip}
is writable and will remain set until cleared by software.

Just as for machine level, the taking of interrupt traps at supervisor
level remains essentially the same as specified by the {\RISCV}
Privileged Architecture.
An interrupt traps into \mbox{S-mode} (or \mbox{HS-mode}) whenever all
of the following are true:
(a)~either the current privilege mode is \mbox{S-mode} and
supervisor-level interrupts are enabled by the SIE bit of \z{sstatus},
or the current privilege mode has less privilege than \mbox{S-mode};
(b)~matching bits in \z{sip} and \z{sie} are both
one, or, if the hypervisor extension is implemented, matching bits in
\z{hip} and \z{hie} are both one; and
(c)~if the hypervisor extension is implemented, the corresponding bit
in \z{hideleg} is zero.

@@

For a given interrupt number, if the corresponding bit in
\z{sie} is read-only zero, then the interrupt's priority
number in the supervisor-level \z{iprio} array must be read-only zero
as well.
The priority number for a supervisor-level external interrupt
(bits 15:8 of \z{iprio2}) must also be read-only zero.
Aside from these two restrictions, implementations may freely choose
which priority number fields are settable and which are read-only
zeros.

@@

The value of \z{stopi} is zero unless:
(a)~there is an interrupt that is both pending in \z{sip}
and enabled in \z{sie}, or, if the hypervisor extension is
implemented, both pending in \z{hip} and enabled in \z{hie}; and
(b)~the interrupt is not delegated to a lower privilege level
(by \z{hideleg}, if the hypervisor extension is implemented).
When there is a pending-and-enabled major interrupt for supervisor
level, field IID is the major identity number of the highest-priority
interrupt, and field IPRIO indicates its priority.

### 2023-05-09 doc/src/preface.tex
Allocate local interrupts for RAS events
Signed-off-by: John Hauser <31252952+jhauser-us@users.noreply.github.com>
Change descriptions of high-half CSRs to match latest Priv ISA
Signed-off-by: John Hauser <31252952+jhauser-us@users.noreply.github.com>

This document describes an Advanced Interrupt Architecture
being proposed for {\RISCV} systems.

@@

\section*{Changes for RC3 (Ratification Candidate 3)}

@@

>>>>

This document describes an Advanced Interrupt Architecture
for {\RISCV} systems.

@@

\section*{Changes for RC4 (Ratification Candidate 4)}

Changed the description (but not the actual function)
of \emph{high-half} CSRs and their partner CSRs
to match the latest {\RISCV} Privileged ISA specification.
(An example of a high-half CSR is \z{miph},
and its partner here is \z{mip}.)

\section*{Changes for RC3}

@@

Allocated major interrupts 35 and 43 for signaling RAS events
(Section~\ref{sec:majorIntrs}).
